Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A study based on the parametric g-formula

### 2024-04-10 output/paper/manuscript.qmd
Shorter abstract and discussion.

embed-resources: true
format:
  docx:
    page-width: 8.3
    citeproc: true
    mermaid:
      theme: neutral
  pdf:
    documentclass: article
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    pdf-engine: lualatex
    number-sections: true
    toc: false
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{lineno}
        \usepackage{rotating}
        \usepackage{float}
        \usepackage{typearea}
    include-before-body:
      text: |
        \linenumbers

@@

Evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of inattention from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}, indicating lower attention. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattention in children and with the homeostasis of the {{< acr hpa >}} axis.

@@

To the best of our knowledge, no other study has investigated the effects of childhood exposure to multiple classes of non-persistent {{< acr edcs >}} in relation to attentional function.
More generally, the literature on childhood exposure to non-persistent {{< acr edcs >}} and other neurodevelopment outcomes in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @LiPapandonatosCalafat:2019; @BalalianWhyattLiu:2019; @ShoaffCoullWeuve:2020; @JankowskaPolanskaHanke:2019; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019].
González-Alzaga et al. and Cartier et al. evaluated cross-sectional associations between {{< acr dap >}} metabolites and subtests of the Wechsler Intelligence Scale for Children [@WechslerKodama:2003] in European children with ages between 6 and 11 years. Higher levels of {{< acr dap >}} metabolites (DMP, DMTP, DMDTP, DEP, DETP, and DEDTP) were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites (DEP, DETP, DEDTP) were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016].
There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. Specifically, Bouchard et al. found evidence of a cross-sectional association between dimethyl alkylphosphate metabolites (DMP, DMTP, and DMDTP) and {{< acr adhd >}} in children aged 8 to 15 years from {{< acr nhanes >}}, while Yu et al. found a dose-response relationship between DMP and {{< acr adhd >}} in Taiwanese children aged 4 to 15 years.

>>>>

embed-resources: true
format:
  docx:
    page-width: 8.3
    citeproc: true
    mermaid:
      theme: neutral
    output-file: "Main document - anonymous"
  pdf:
    documentclass: article
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    pdf-engine: lualatex
    number-sections: true
    toc: false
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{lineno}
        \usepackage{rotating}
        \usepackage{float}
        \usepackage{typearea}
    include-before-body:
      text: |
        \linenumbers

@@

Evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate associations between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of inattention, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}, indicating lower attention. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattention and with the homeostasis of the {{< acr hpa >}} axis.

@@

To the best of our knowledge, no other study has investigated the effects of childhood exposure to multiple classes of non-persistent {{< acr edcs >}} in relation to attentional function.
More generally, the literature on childhood exposure to non-persistent {{< acr edcs >}} and other neurodevelopment outcomes in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @LiPapandonatosCalafat:2019; @BalalianWhyattLiu:2019; @ShoaffCoullWeuve:2020; @JankowskaPolanskaHanke:2019; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019].

In children aged 6 to 11 years, higher levels of {{< acr dap >}} metabolites were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016]. There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016].
